---
title: "NYPD Shooting Incident Analysis"
author: "by Susan Qu"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
library(readr)
library(dplyr)
library(ggplot2)
library(lubridate)
library(scales)
library(broom)
```

```{r data-import}
# Import and clean data
nypd_data <- read_csv("NYPD_Shooting_Incident_Data__Historic_.csv")
nypd_clean <- nypd_data %>%
  mutate(
    OCCUR_DATE = mdy(OCCUR_DATE),
    Year = year(OCCUR_DATE),
    Month = month(OCCUR_DATE, label = TRUE),
    MonthNum = month(OCCUR_DATE),
    Season = case_when(
      MonthNum %in% c(12, 1, 2) ~ "Winter",
      MonthNum %in% c(3, 4, 5) ~ "Spring",
      MonthNum %in% c(6, 7, 8) ~ "Summer",
      MonthNum %in% c(9, 10, 11) ~ "Fall"
    )
  ) %>%
  filter(!is.na(OCCUR_DATE))
```

## Executive Summary

This report analyzes NYPD shooting incident data across time periods to identify patterns and trends. It includes:

- Borough Analysis
  - Shooting incidents by Borough
  - Borough trends over time
- Incidens Overall by Year and Month
- Time of Day Analysis
- Predictive Model: Factors Contributing to Shooting Incidents
- Key Findings
- Potential Biases


## Borough Analysis

```{r borough-analysis, fig.height=6}
# Clean borough data and create summary
borough_summary <- nypd_clean %>%
  filter(!is.na(BORO) & BORO != "") %>%
  group_by(BORO) %>%
  summarise(incidents = n(), .groups = 'drop') %>%
  mutate(
    percentage = round(incidents/sum(incidents)*100, 1),
    BORO = case_when(
      BORO == "BRONX" ~ "Bronx",
      BORO == "BROOKLYN" ~ "Brooklyn",
      BORO == "MANHATTAN" ~ "Manhattan",
      BORO == "QUEENS" ~ "Queens",
      BORO == "STATEN ISLAND" ~ "Staten Island",
      TRUE ~ BORO
    )
  ) %>%
  arrange(desc(incidents))

# Borough bar chart
ggplot(borough_summary, aes(x = reorder(BORO, incidents), y = incidents, fill = BORO)) +
  geom_col() +
  geom_text(aes(label = paste0(incidents, "\n(", percentage, "%)")),
            hjust = -0.1, color = "black", fontface = "bold") +
  coord_flip() +
  labs(title = "Shooting Incidents by Borough",
       subtitle = "Total incidents across all years",
       x = "Borough", y = "Number of Incidents") +
  scale_fill_viridis_d() +
  theme_minimal() +
  theme(legend.position = "none")
```

## Borough Trends Over Time

```{r borough-trends, fig.height=6}
# Borough trends over time
borough_yearly <- nypd_clean %>%
  filter(!is.na(BORO) & BORO != "") %>%
  mutate(
    BORO = case_when(
      BORO == "BRONX" ~ "Bronx",
      BORO == "BROOKLYN" ~ "Brooklyn",
      BORO == "MANHATTAN" ~ "Manhattan",
      BORO == "QUEENS" ~ "Queens",
      BORO == "STATEN ISLAND" ~ "Staten Island",
      TRUE ~ BORO
    )
  ) %>%
  group_by(Year, BORO) %>%
  summarise(incidents = n(), .groups = 'drop')

ggplot(borough_yearly, aes(x = Year, y = incidents, color = BORO)) +
  geom_line(size = 1.2) +
  geom_point(size = 2) +
  labs(title = "Shooting Incidents by Borough Over Time",
       subtitle = "Yearly trends across NYC boroughs",
       x = "Year", y = "Number of Incidents",
       color = "Borough") +
  scale_color_viridis_d() +
  theme_minimal() +
  theme(legend.position = "bottom")
```

## Heatmap of incidents overall by year and month
```{r monthly-trends-by-year-month, fig.height=5}
yearly_monthly <- nypd_clean %>%
  group_by(Year, MonthNum) %>%
  summarise(incidents = n(), .groups = 'drop')

ggplot(yearly_monthly, aes(x = MonthNum, y = factor(Year), fill = incidents)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkred", name = "Incidents") +
  scale_x_continuous(breaks = 1:12, labels = month.abb) +
  labs(title = "Shooting Incidents Heatmap: Year vs Month",
       x = "Month", y = "Year") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Time of Day Analysis

```{r time-of-day-analysis, fig.height=8}
# Prepare time of day data
time_data <- nypd_clean %>%
  filter(!is.na(OCCUR_TIME)) %>%
  mutate(
    # Extract hour from time
    Hour = hour(hms(OCCUR_TIME)),
    # Create time periods
    TimeOfDay = case_when(
      Hour >= 6 & Hour < 12 ~ "Morning",
      Hour >= 12 & Hour < 18 ~ "Afternoon",
      Hour >= 18 & Hour < 24 ~ "Evening",
      TRUE ~ "Night"
    )
  )

# 1. Hourly distribution
hourly_incidents <- time_data %>%
  group_by(Hour) %>%
  summarise(incidents = n(), .groups = 'drop')

p_hourly <- ggplot(hourly_incidents, aes(x = Hour, y = incidents)) +
  geom_col(fill = "steelblue", alpha = 0.8) +
  geom_smooth(method = "loess", color = "red", se = FALSE) +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  labs(title = "Shooting Incidents by Hour of Day",
       subtitle = "24-hour distribution with trend line",
       x = "Hour of Day", y = "Number of Incidents") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_hourly)
```


## Predictive Model: Factors Contributing to Shooting Incidents

```{r model-data-prep}
# Prepare data for modeling
model_data <- nypd_clean %>%
  filter(!is.na(BORO) & BORO != "" & !is.na(OCCUR_TIME)) %>%
  mutate(
    # Clean borough names
    BORO = case_when(
      BORO == "BRONX" ~ "Bronx",
      BORO == "BROOKLYN" ~ "Brooklyn",
      BORO == "MANHATTAN" ~ "Manhattan",
      BORO == "QUEENS" ~ "Queens",
      BORO == "STATEN ISLAND" ~ "Staten Island",
      TRUE ~ BORO
    ),
    # Extract hour from time
    Hour = hour(hms(OCCUR_TIME)),
    # Create time periods
    TimeOfDay = case_when(
      Hour >= 6 & Hour < 12 ~ "Morning",
      Hour >= 12 & Hour < 18 ~ "Afternoon",
      Hour >= 18 & Hour < 24 ~ "Evening",
      TRUE ~ "Night"
    ),
    # Create weekend indicator
    Weekday = wday(OCCUR_DATE, label = TRUE),
    IsWeekend = ifelse(Weekday %in% c("Sat", "Sun"), 1, 0),
    # Create summer indicator (higher crime months)
    IsSummer = ifelse(Season == "Summer", 1, 0)
  ) %>%
  # Group by relevant factors and count incidents
  group_by(Year, BORO, Season, TimeOfDay, IsWeekend, IsSummer, MonthNum) %>%
  summarise(IncidentCount = n(), .groups = 'drop')

# Display summary statistics
cat("Model Data Summary:\n")
summary(model_data)
```

```{r linear-model}
# Build linear regression model
incident_model <- lm(IncidentCount ~ BORO + Season + TimeOfDay + IsWeekend +
                     IsSummer + MonthNum + Year, data = model_data)

# Model summary
summary(incident_model)

# Extract and display coefficients
model_coef <- data.frame(
  Factor = names(coef(incident_model)),
  Coefficient = coef(incident_model),
  row.names = NULL
) %>%
  filter(Factor != "(Intercept)") %>%
  arrange(desc(abs(Coefficient)))

print(model_coef)
```

```{r model-visualization, fig.height=8}

# Get tidy model results
tidy_model <- tidy(incident_model, conf.int = TRUE) %>%
  filter(term != "(Intercept)") %>%
  mutate(
    term = gsub("BORO|Season|TimeOfDay", "", term),
    significant = ifelse(p.value < 0.05, "Significant", "Not Significant")
  )

# Coefficient plot
ggplot(tidy_model, aes(x = reorder(term, estimate), y = estimate,
                       color = significant)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.7) +
  coord_flip() +
  labs(title = "Factors Contributing to Shooting Incidents",
       subtitle = "Linear regression coefficients with 95% confidence intervals",
       x = "Factors", y = "Coefficient Estimate",
       color = "Significance (p < 0.05)") +
  scale_color_manual(values = c("Significant" = "red", "Not Significant" = "gray")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r feature-importance, fig.height=6}
# Create feature importance plot based on absolute coefficient values
importance_data <- tidy_model %>%
  mutate(
    abs_estimate = abs(estimate),
    term_clean = case_when(
      grepl("Brooklyn|Bronx|Manhattan|Queens", term) ~ paste("Borough:", term),
      grepl("Fall|Spring|Summer|Winter", term) ~ paste("Season:", term),
      grepl("Evening|Morning|Night", term) ~ paste("Time:", term),
      term == "IsWeekend" ~ "Weekend",
      term == "IsSummer" ~ "Summer Period",
      term == "MonthNum" ~ "Month Number",
      term == "Year" ~ "Year Trend",
      TRUE ~ term
    )
  ) %>%
  top_n(10, abs_estimate)

ggplot(importance_data, aes(x = reorder(term_clean, abs_estimate),
                           y = abs_estimate, fill = significant)) +
  geom_col() +
  coord_flip() +
  labs(title = "Top 10 Most Important Factors",
       subtitle = "Based on absolute coefficient values",
       x = "Factors", y = "Absolute Coefficient Value",
       fill = "Significance") +
  scale_fill_manual(values = c("Significant" = "darkred", "Not Significant" = "lightgray")) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

```{r time-heatmap, fig.height=6}
# 3. Heatmap: Hour vs Day of Week
time_weekday <- time_data %>%
  mutate(
    Weekday = wday(OCCUR_DATE, label = TRUE, abbr = FALSE),
    Hour = hour(hms(OCCUR_TIME))
  ) %>%
  group_by(Weekday, Hour) %>%
  summarise(incidents = n(), .groups = 'drop')

ggplot(time_weekday, aes(x = Hour, y = Weekday, fill = incidents)) +
  geom_tile() +
  scale_fill_gradient(low = "lightblue", high = "darkred", name = "Incidents") +
  scale_x_continuous(breaks = seq(0, 23, 2)) +
  labs(title = "Shooting Incidents Heatmap: Hour vs Day of Week",
       subtitle = "Darker colors indicate higher incident counts",
       x = "Hour of Day", y = "Day of Week") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Key Findings

The linear regression model identifies several significant factors contributing to shooting incidents:

1. **Borough Effects**: Different boroughs show varying baseline incident rates; for example, Brooklyn has the highest number of incidents, while Staten Island has the lowest.
2. **Seasonal Patterns**: Summer months typically see higher incident rates; incidents spike in July and August.
3. **Time of Day**: Evening and night hours show different patterns than daytime; incidents are more frequent after 6 PM, and peaks are observed around midnight.
4. **Temporal Trends**: Year-over-year changes in incident patterns; recent years have shown a decline in incidents.
5. **Weekend Effect**: Weekends may have different incident rates than weekdays; Saturday nights are particularly notable.

## Potential Biases

### **Reporting Bias**
- **Under-reporting**: Not all shooting incidents may be reported to police, particularly in communities with distrust of law enforcement. See [Economic and Policy Insights](https://www.osc.ny.gov/reports/concerning-growth-hate-crime-new-york-state#:~:text=6,victims'%20ability%20to%20report%20crimes).

### **Population Bias**
- **Demographic differences**: From the analysis, while Staten Island has the lowest number of incidents, it is important to consider that Staten Island has a significantly smaller population compared to other boroughs. This can skew the perception of safety and incident rates. Brooklyn is the most populous of New York City's five boroughs, while Staten Island is the least populated. Brooklyn has over 2.6 million residents, whereas Staten Island has less than 500,000. See [NYC Population Data](https://www1.nyc.gov/site/planning/data-maps/nyc-population/population-facts.page) for more details.

## References
- NYPD Shooting Incident Data: [NYPD Data Portal](https://data.cityofnewyork.us/Public-Safety/NYPD-Shooting-Incident-Data-Historic-/qgea-i56i)
- NYC Population Data: [NYC Planning](https://www1.nyc.gov/site/planning/data-maps/nyc-population/population-facts.page)
- Economic and Policy Insights: [Hate Crime Report](https://www.osc.ny.gov/reports/concerning-growth-hate-crime-new-york-state#:~:text=6,victims'%20ability%20to%20report%20crimes)

## Github Link:
- repo: https://github.com/SusanQu/boulder_msds_ds_as_a_field
